Geometric models reveal behavioral and neural signatures of transforming experiences into memories

### 2020-10-16 paper/main.tex
minor typo fixes in main document, regenerated diffs

Chen et al. (2017) collected hand-annotated information about each of 1000 (manually delineated) time segments spanning the roughly 50 minute video used in their study~\cite{ChenEtal17}.  Each annotation included: a brief narrative description of what was happening, the location where the action took place, the names of any characters on the screen, and other similar details (for a full list of annotated features, see \textit{Methods}).  We took the union of all unique words (excluding stop words, such as ``and,'' ``or,'' ``but,'' etc.) across all features from all annotations as the vocabulary for the topic model.  We then concatenated the sets of words across all features contained in overlapping sliding windows of (up to) 50 annotations, and treated each window as a single document for the purpose of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 32 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the episode (see \textit{Methods}; Fig.~\ref{fig:schematic}, Supp.\ Fig.~\topics).  We note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06} in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize how the properties of collections of documents change over time, our sliding window approach allows us to examine the topic dynamics within a single document (or video).  Specifically, our approach yielded (via the topic proportions matrix) a single ``topic vector'' for each sliding window of annotations transformed by the topic model.  We then stretched (interpolated) the resulting windows-by-topics matrix to match the time series of the 1,976 fMRI volumes collected as participants viewed the episode.

@@

Clustering scores are often used by memory researchers to characterize how people organize their memories of words on a studied list~\citep{PolyEtal09}.  We defined analogous measures to characterize how participants organized their memories for episodic events (see \textit{Methods} for details).  Temporal clustering refers to the extent to which participants group their recall responses according to encoding position.  Overall, we found that sequentially viewed episode events tended to appear nearby in participants' recall event sequences (mean clustering score: 0.732, SEM: 0.033).  Participants with higher temporal clustering scores tended to exhibit better overall memory for the episode, according to both Chen et al. (2017)~\cite{ChenEtal17}'s hand-counted numbers of recalled scenes from the episode (Pearson's $r(15) = 0.49,~p = 0.046,~95\%~\mathrm{CI} = [0.25, 0.76]$) and the numbers of episode events that best-matched at least one recall event (i.e., model-estimated number of events recalled; Pearson's $r(15) = 0.59,~p = 0.013,~95\%~\mathrm{CI} = [0.31, 0.80]$).  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity~\citep{MannKaha12}.  We found that participants tended to recall semantically similar episode events together (mean clustering score: 0.650, SEM: 0.032), and that semantic clustering score was also related to both hand-counted  (Pearson's $r(15) = 0.65,~p = 0.004,~95\%~\mathrm{CI} = [0.31, 0.85]$) and model-estimated (Pearson's $r(15) = 0.58,~p = 0.015,~95\%~\mathrm{CI} = [0.10, 0.83]$) numbers of recalled events.

@@

Distinctiveness is intended to capture the ``specificity'' of recall.  In other words, distinctiveness quantifies the extent to which a given recall event reflects the most similar episode event over and above other episode events.  Intuitively, distinctiveness is like a normalized variant of our precision metric.  Whereas precision solely measures how much detail about an episode was captured in someone's recall, distinctiveness penalizes details that also pertain to other episode events.  We define the distinctiveness of an event's recall as its precision expressed in standard deviation units with respect to other episode events.  Specifically, for a given recall event, we compute the correlation between its topic vector and that of each episode event.  This yields a distribution of correlation coefficients (one per episode event).  We subtract the mean and divide by the standard deviation of this distribution to $z$-score the coefficients.  The maximum value in this distribution (which, by definition, belongs to the episode event that best matches the given recall event) is that recall event's distinctiveness score.  In this way, recall events that match one episode event far better than all other episode events will receive a high distinctiveness score.  By contrast, a recall event that matches all episode events roughly equally will receive a comparatively low distinctiveness score.

@@

Figure \ref{fig:precision-detail} also illustrates the differences between high and low distinctiveness scores.  We extracted the set of sentences comprising the most distinctive recall event (P9) and least distinctive recall event (P6) corresponding to the example episode event shown in Panel C (event 21).  We also extracted the annotations for all episode events whose content these participants' single recall events described.  We assigned each episode event a unique color (Fig.~\ref{fig:precision-detail}E), and colored each recalled sentence (Panel F) according to the episode events they best matched.  Visual inspection of Panel F reveals that the most distinctive recall's content is tightly concentrated around event 21, whereas the least distinctive recall incorporates content from a much wider range of episode events.

@@

In addition to enabling us to visualize the episode's high-level essence, describing the episode as a geometric trajectory also enables us to drill down to individual words and quantify how each word relates to the memorability of each event.  This provides another approach to examining participants' recall for low-level details beyond the precision and distinctiveness measures we defined above.  The results displayed in Figures \ref{fig:list-learning}C and \ref{fig:precision-detail}A suggest that certain events were remembered better than others.  Given this, we next asked asked whether the events that were generally remembered precisely or imprecisely tended to reflect particular content.  Because our analysis framework projects the dynamic episode content and participants' recalls into a shared space, and because the dimensions of that space represent topics (which are, in turn, sets of weights over known words in the vocabulary), we are able to recover the weighted combination of words that make up any point (i.e., topic vector) in this space.  We first computed the average precision with which participants recalled each of the 30 episode events (Fig.~\ref{fig:topics}A; note that this result is analogous to a serial position curve created from our precision metric).  We then computed a weighted average of the topic vectors for each episode event, where the weights reflected how precisely each event was recalled.  To visualize the result, we created a ``wordle'' image~\citep{MuelEtal18} where words weighted more heavily by more precisely-remembered topics appear in a larger font (Fig.~\ref{fig:topics}B, green box).  Across the full episode, content that weighted heavily on topics and words central to the major foci of the episode (e.g., the names of the two main characters, ``Sherlock'' and ``John,'' and the address of a major recurring location, ``221B Baker Street'') was best remembered.  An analogous analysis revealed which themes were less-precisely remembered.  Here, in computing the weighted average over events' topic vectors, we weighted each event in inverse proportion to its average precision (Fig.~\ref{fig:topics}B, red box).  The least precisely remembered episode content reflected information that was extraneous to the episode's essence, such as the proper names of relatively minor characters (e.g., ``Mike,'' ``Molly,'' and ``Lestrade'') and locations (e.g., ``St. Bartholomew's Hospital'').

>>>>

Chen et al. (2017) collected hand-annotated information about each of 1000 (manually delineated) time segments spanning the roughly 50 minute video used in their study~\cite{ChenEtal17}.  Each annotation included: a brief narrative description of what was happening, the location where the action took place, the names of any characters on the screen, and other similar details (for a full list of annotated features, see \textit{Methods}).  We took the union of all unique words (excluding stop words, such as ``and,'' ``or,'' ``but,'' etc.) across all features from all annotations as the vocabulary for the topic model.  We then concatenated the sets of words across all features contained in overlapping sliding windows of (up to) 50 annotations, and treated each window as a single document for the purpose of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 32 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the episode (see \textit{Methods}; Fig.~\ref{fig:schematic}, Supp.\ Fig.~\topics).  We note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06} in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize how the properties of collections of documents change over time, our sliding window approach allows us to examine the topic dynamics within a single document (or video).  Specifically, our approach yielded (via the topic proportions matrix) a single ``topic vector'' for each sliding window of annotations transformed by the topic model.  We then stretched (interpolated) the resulting windows-by-topics matrix to match the time series of the 1976 fMRI volumes collected as participants viewed the episode.

@@

Clustering scores are often used by memory researchers to characterize how people organize their memories of words on a studied list~\citep{PolyEtal09}.  We defined analogous measures to characterize how participants organized their memories for episodic events (see \textit{Methods} for details).  Temporal clustering refers to the extent to which participants group their recall responses according to encoding position.  Overall, we found that sequentially viewed episode events tended to appear nearby in participants' recall event sequences (mean clustering score: 0.732, SEM: 0.033).  Participants with higher temporal clustering scores tended to exhibit better overall memory for the episode, according to both Chen et al. (2017)~\cite{ChenEtal17}'s hand-counted numbers of recalled scenes from the episode (Pearson's $r(15) = 0.49,~p = 0.046,~95\%~\mathrm{CI} = [0.25, 0.76]$) and the numbers of episode events that best-matched at least one recall event (i.e., model-estimated number of events recalled; Pearson's $r(15) = 0.59,~p = 0.013,~95\%~\mathrm{CI} = [0.31, 0.80]$).  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity~\citep{MannKaha12}.  We found that participants tended to recall semantically similar episode events together (mean clustering score: 0.650, SEM: 0.032), and that semantic clustering scores were also related to both hand-counted  (Pearson's $r(15) = 0.65,~p = 0.004,~95\%~\mathrm{CI} = [0.31, 0.85]$) and model-estimated (Pearson's $r(15) = 0.58,~p = 0.015,~95\%~\mathrm{CI} = [0.10, 0.83]$) numbers of recalled events.

@@

Distinctiveness is intended to capture the ``specificity'' of recall.  In other words, distinctiveness quantifies the extent to which a given recall event reflects the most similar episode event over and above other episode events.  Intuitively, distinctiveness is like a normalized variant of our precision metric.  Whereas precision solely measures how much detail about an event was captured in someone's recall, distinctiveness penalizes details that also pertain to other episode events.  We define the distinctiveness of an event's recall as its precision expressed in standard deviation units with respect to other episode events.  Specifically, for a given recall event, we compute the correlation between its topic vector and that of each episode event.  This yields a distribution of correlation coefficients (one per episode event).  We subtract the mean and divide by the standard deviation of this distribution to $z$-score the coefficients.  The maximum value in this distribution (which, by definition, belongs to the episode event that best matches the given recall event) is that recall event's distinctiveness score.  In this way, recall events that match one episode event far better than all other episode events will receive a high distinctiveness score.  By contrast, a recall event that matches all episode events roughly equally will receive a comparatively low distinctiveness score.

@@

Figure \ref{fig:precision-detail} also illustrates the differences between high and low distinctiveness scores.  We extracted the set of sentences comprising the most distinctive recall event (P9) and least distinctive recall event (P6) corresponding to the example episode event shown in Panel C (event 21).  We also extracted the annotations for all episode events whose content these participants' single recall events touched on.  We assigned each episode event a unique color (Fig.~\ref{fig:precision-detail}E), and colored each recalled sentence (Panel F) according to the episode events they best matched.  Visual inspection of Panel F reveals that the most distinctive recall's content is tightly concentrated around event 21, whereas the least distinctive recall incorporates content from a much wider range of episode events.

@@

In addition to enabling us to visualize the episode's high-level essence, describing the episode as a geometric trajectory also enables us to drill down to individual words and quantify how each word relates to the memorability of each event.  This provides another approach to examining participants' recall for low-level details beyond the precision and distinctiveness measures we defined above.  The results displayed in Figures \ref{fig:list-learning}C and \ref{fig:precision-detail}A suggest that certain events were remembered better than others.  Given this, we next asked whether the events that were generally remembered precisely or imprecisely tended to reflect particular content.  Because our analysis framework projects the dynamic episode content and participants' recalls into a shared space, and because the dimensions of that space represent topics (which are, in turn, sets of weights over known words in the vocabulary), we are able to recover the weighted combination of words that make up any point (i.e., topic vector) in this space.  We first computed the average precision with which participants recalled each of the 30 episode events (Fig.~\ref{fig:topics}A; note that this result is analogous to a serial position curve created from our precision metric).  We then computed a weighted average of the topic vectors for each episode event, where the weights reflected how precisely each event was recalled.  To visualize the result, we created a ``wordle'' image~\citep{MuelEtal18} where words weighted more heavily by more precisely remembered topics appear in a larger font (Fig.~\ref{fig:topics}B, green box).  Across the full episode, content that weighted heavily on topics and words central to the major foci of the episode (e.g., the names of the two main characters, ``Sherlock'' and ``John,'' and the address of a major recurring location, ``221B Baker Street'') was best remembered.  An analogous analysis revealed which themes were less-precisely remembered.  Here, in computing the weighted average over events' topic vectors, we weighted each event in inverse proportion to its average precision (Fig.~\ref{fig:topics}B, red box).  The least precisely remembered episode content reflected information that was extraneous to the episode's essence, such as the proper names of relatively minor characters (e.g., ``Mike,'' ``Molly,'' and ``Lestrade'') and locations (e.g., ``St. Bartholomew's Hospital'').
